\sys: Light-touch Data-Parallel Shell Processing

### 2020-05-17 pldi/draft.tex
Minor intro corrections

@@

The second group is \emph{shell users}, who use POSIX shell constructs to combine multiple such commands from many languages into their scripts and are thus left with only a few options for incorporating parallelism (see~\S\ref{related}). \kk{I don't really like the reference to such a later section. Can we remove it?}
% If versed into parallelism, 
% Users versed in computing can apply 
% these tools are command-unaware thus at risk of breaking program semantics,
% require manual changes to the code to make parallelism explicit,
% at times leverage command flags like the ones mentioned earlier---but each command 
% oblivious? agnostic?  
One option is manual tools such GNU \ttt{parallel}~\cite{Tange2011a}, \ttt{ts}~\cite{tsp}, \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm};
  these tools are either command-unaware thus at risk of breaking program semantics, or too coarse-grained \kk{i.e. only being capable to exploit parallelism in the level of entire scripts} thus applying at the level of entire scripts rather than individual components.
% Tools will attempt to parallelize anything 
% use the composition abstractions of the \unix shell, 
% and are often prototypical in nature---focusing on quickly prototyping a solution for the task at hand rather building an optimal solution to a recurring pattern.
% short lived, changing quickly
% (ii) ad-hoc flags provided by command developers and mentioned above.
Another option is using the shell's concurrency primitives (\eg \ttt{\&}, \ttt{wait}, \ttt{for}) \kk{Is \ttt{for} a concurrency primitive?} % for explicit parallelism, 
 that come at a cost of manual effort to carefully split inputs, manually rewrite scripts, and explicitly orchestrate execution---an expensive and error-prone process.
% re-writing of the entire script from scratch; unfortunately, this process is expensive and can introduce new bugs, cascading changes, and divergence from legacy functionality---but also foregoes all the productivity benefits of the original script.
% The fact that a programmer often belongs to both groups at the same time---as composed scripts can themselves form a command used in other scrips---compounds the challenge.
And\todo{/To top it off} all these options assume a good understanding of parallelism;
  users whose domain of expertise falls outside computing---\todo{routinely} writing shell scripts to accomplish their tasks---are left without options.% automation when it comes to parallelization.

To address this challenge, we develop \sys, a shell parallelizing\todo{/that parallelizes} POSIX shell scripts.
\kk{Mention that it outputs shell scripts.} % to their parallel equivalents.
% \sys offers both classes of users in the second group such much-needed automation.
\sys benefits both programmer groups mentioned earlier, with a particular emphasis on users.
Command developers are provided a set of abstractions, akin to lightweight type annotations, for expressing the parallelizability properties of their commands:
  rather than providing a command's full observable behavior, these annotations focus primarily on their interaction with state.
Shell users, on the other hand, are provided full automation:
  \sys extracts parallelism latent in their scripts by analyzing their parallelizability properties. % of their fragments.
% benefiting scripting experts and non-experts alike by offering automated POSIX-aware parallelization of shell-scripts as long as command developers
\sys's transformations are conservative, in that they do not attempt to parallelize fragments that lack sufficient information---\ie at worst, \sys will choose to not improve performance rather than risking breakage. \kk{Should we mention that we only show this empirically?}

@@

These components are finally paired with \sys's runtime component.
% \sys's runtime  component additional packs
% Aside from these two components, \sys packs a runtime component that serves dual purpose:
Aware of the \unix philosophy and abstractions, it packs a small library of highly-optimized data combiners\todo{/combinators} as well as high-performance primitives for eager data splitting and merging.
These address several practical challenges such as task parallelism, deadlocks, and performance bottlenecks\kk{I like the flow of this sentence but I am not sure it makes a lot of sense to name these challenges here.}, and were developed by uncovering several pathological situations, on many of which we report.

@@

We evaluate \sys using
  (i) a series of benchmarks, ranging from classic one-liners found in the \unix literature to modern data-processing pipelines;
  (ii) three large and complex use cases, covering temperature analysis, web crawling and indexing, and container orchestration; and
  (iii) a series of targeted micro-benchmarks inspired by (and comparing with) possible alternatives.
The results demonstrate significant advantages in correctness, manual effort, and performance---with \sys's transformations offering an order-of-magnitude performance improvements aided further by \sys's optimized runtime components. \kk{I would remove the ``aided further...'' part.}
Naturally, improvements on individual commands depend on several factors such as the script's execution time, the pipeline depth, and the parallelizability class under which a command falls. \kk{Do we need to mention this last sentence here? Without context it is pretty difficult for anyone to get something out of it.}

@@

The paper is structured as follows.
It starts by introducing the necessary background on shell scripting and overviewing \sys~\sx{bg}.
Sections \ref{distributability}--\ref{impl} highlight key contributions:
\begin{itemize}
  \item
  \S\ref{distributability}
    studies the parallelizability properties of a large class of shell commands and introduces a lightweight annotation language for commands that can be executed in a data-parallel manner.

@@

  \item
  \S\ref{impl} describes \sys's implementation, discussing several challenges related to \sys's translation, optimization, and runtime phases.
% \kk{Rewrite that when we know for sure what the section contains.}

@@

\heading{Command Inputs and Outputs}

@@

\heading{Command-line Options}
As described earlier~\sx{bg}, com\-mand-line options and environment
variables can affect the operation of a command.  Options are a
particularly popular way to control a command, directly affecting a
command's parallelizability classification.  That is, commands are
assigned to default classes, and options can change the class in which
they belong.  As a simple example, consider \ttt{cat}.  By default
\ttt{cat} is in \sta, but with \ttt{-n} it jumps into \pur because it
has to keep track of a counter and output the line number with each line.

@@

%% \heading{Command Inputs and Outputs}
%% A second concern captured by the AL is the order of inputs and outputs.
%% Commands generally read multiple input streams (and, at cases, write to multiple output streams).
%% The invariants between inputs and outputs must be maintained over the distributed execution of the program.
%% % but these maintain different invariants, enabling very different distributability characteristics.

@@

\heading{Parallelizability Annotation Language}

@@

\kk{We could add the grammar of PAL in an appendix if we want}

@@

\heading{Parallelizable Pure Commands}

@@

\heading{Edges -- Files}

@@

\heading{Nodes -- Commands}

@@

\heading{\todo{One-input-output} Commands}

@@

\heading{Stateless and Parallelizable Pure Commands}

@@

\heading{Parallelization Transformations}
%
Based on these equations, we can define a \emph{node parallelization
  transformation} on a stateless node $v$ that is preceded by a
\ttt{cat} with $n$ input streams and is followed by a node $v'$. The
transformation replaces $v$ with $n$ new nodes, routing each of the
$n$ input streams to one of them, and commutes the \ttt{cat} node
after them to concatenate their outputs and transfer them to
$v'$. Since each incoming edge represents a stream of data $x_i :
D^*$, and the only behaviour of a dataflow graph is its output, this
optimization $ v(x_1.x_2...x_n, s) \Rightarrow v(x_1, s).v(x_2,
s)...v(x_n, s)$ can be shown to preserve the behaviour of the
graph. Furthermore, it can be straightforwardly extended to pure
commands that can be implemented by a map-fold pair $(m, r)$ as $
v(x_1.x_2...x_n, s) \Rightarrow r(m(x_1, s).m(x_2, s)...m(x_n, s),
s)$. A visual representation of this transformation can be seen in
\Cref{fig:parallelization-transformation}.

@@

\heading{Parallelizable Regions}
%
%% \kk{The goal of this subsection is to explain that Dish does a safe
%%   but effective analysis to identify which parts of the scripts to
%%   parallelize.}
%
Parallelizable regions are program sub-expressions that can be
parallelized safely without altering the total program behaviour.
%% respect to their sequential execution.
The search for these regions can be guided by the structure of shells
programs, which inherently contains information about components that
can be executed independently and barriers that are natural points for
enforcing synchronization. Consider the following example:

@@

\heading{Translation Pass}

@@

\heading{Pure Fold Function Implementations}

@@

    \begin{subfigure}[b]{0.24\textwidth}
        \centering
        \includegraphics[width=\textwidth]{\detokenize{./figs/topn_throughput_scaleup.pdf}}
        %% \caption{TODO}
        %% \label{eval:topn}
    \end{subfigure}%
    ~
    \begin{subfigure}[b]{0.24\textwidth}
        \centering
        \includegraphics[width=\textwidth]{\detokenize{./figs/wf_throughput_scaleup.pdf}}
        %% \caption{TODO}
        %% \label{eval:wf}
    \end{subfigure}%
    ~
    \begin{subfigure}[b]{0.24\textwidth}
        \centering
        \includegraphics[width=\textwidth]{\detokenize{./figs/shortest_scripts_throughput_scaleup.pdf}}
        %% \caption{TODO}
        %% \label{eval:spell}
    \end{subfigure}%
    ~
    \begin{subfigure}[b]{0.24\textwidth}
      \centering
      \includegraphics[width=\textwidth]{\detokenize{./figs/alt_bigrams_throughput_scaleup.pdf}}
        %% \caption{TODO}
        %% \label{eval:minimal_sort}
    \end{subfigure}%
    \caption{
      \textbf{Speedup achieved by \sys as a function of the level of parallelism (1--100).} 
      Four different configurations per benchmark:
      (i) \todo{Parallel + Split}, showing the performance of \sys with both eager and split optimizations enabled,
      (ii) \todo{Parallel}, showing the performance of \sys with eager enabled and split disabled,
      (iii) \todo{No Eager}, showing the performance of \sys with split and eager disabled,
      (iv) \todo{Blocking Eager}, showing the performance of \sys with split disabled and the naive blocking eager enabled,
      \nv{Maybe it makes sense to show log x-axis?}
      \kk{How can we also fit two more plots (sort-sort and diff) here? Maybe we can fit one by getting rid of sort (since it also shown in the sort--parallel one). Otherwise we could separate the ``split'' plots from the ones that don't have split.}
    }
    \vspace{-15pt}
    \label{fig:microbenchmarks}
\end{figure*}

\heading{Programs}

\kk{How can we include a result for grep-light?  This is the IO
  intensive CPU-light grep. \sys has no speedup but also no slowdown
  (even if it IO heavy). Should we just mention that in text in
  passing?}

Tab.~\ref{tab:eval} summarizes the collection of micro-benchmarks used to evaluate \sys.
Benchmark names (col. 1) correspond to the names on the plots (Fig.~\ref{fig:microbenchmarks}), and structure (col. 2) summarizes the different classes of commands used in the script.
% The centralized pipelines read files of various sizes from disk and write to disk.
Input sizes and times (col. 3, 4) summarize the characteristics of the largest sequential executions. % (200 files in Fig.~\ref{fig:microbenchmarks}).
Script sizes (col. 5) report on \sys's rewritten output for three different distribution sizes---$1\times$, 10$\times$, and $100\times$.
Size grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
The first number alone is interesting, as it captures \sys's output for the sequential script.
As there is no parallelism involved, it highlights the initial (bare minimum) cost corresponding to using \sys.
% this is the size of 
This change in size does not necessarily translate to observed runtime, as most of the size increase comes from the setup and teardown of FIFOs responsible for interprocess (IPC) communication.
FIFOs are simply pipes made explicit;
  the performance characteristics, including in-kernel buffering and synchronization mechanisms, are identical between the two.

@@

\heading{Annotation-based Distribution}
Recent systems on light-touch distribution, such as
Ignis~\cite{ignis:19} and Mozart~\cite{mozart:19}, are more general
than cluster-computing frameworks.  They also lower (but not entirely
eliminate) developer effort by automating most of the rewriting
process.
% Both work on high-level, managed languages---server-side JavaScript for Ignis and Python for Mozart---that allow runtime transformations without affecting the broader environment under which a program executes.
For the critical points where user input is required, they provide a
high-level declarative DSL with varying (but limited) degrees of
expressive power.
% Annotations are used by ---complicated by the use of third-party
% libraries similar to \sys's extensions.
Unfortunately, a challenge with this approach is that users composing
applications by using third-party libraries do not understand their
distributability details well-enough to annotate them.

>>>>

\nv{Add note about breadth of commands and the impossibility of our goal.}
% https://nextbreakpoint.com/posts/article-compile-code-with-docker.html

@@

The second group is \emph{shell users}, who use POSIX shell constructs to combine multiple such commands from many languages into their scripts and are thus left with only a few options for incorporating parallelism.
% \kk{I don't really like the reference to such a later section. Can we remove it?}
% \nv{We have to say this is not just ``an opinion'', but that we give a scientific treatment/full coverage in the related work section. How far away should the sections be so that you're comfortable referencing to?}
% If versed into parallelism, 
% Users versed in computing can apply 
% these tools are command-unaware thus at risk of breaking program semantics,
% require manual changes to the code to make parallelism explicit,
% at times leverage command flags like the ones mentioned earlier---but each command 
% oblivious? agnostic?  
One option is manual tools such GNU \ttt{parallel}~\cite{Tange2011a}, \ttt{ts}~\cite{tsp}, \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm};
  these tools are either command-unaware thus at risk of breaking program semantics, or too coarse-grained thus only being capable to exploit parallelism at the level of entire scripts rather than individual components.
% Tools will attempt to parallelize anything 
% use the composition abstractions of the \unix shell, 
% and are often prototypical in nature---focusing on quickly prototyping a solution for the task at hand rather building an optimal solution to a recurring pattern.
% short lived, changing quickly
% (ii) ad-hoc flags provided by command developers and mentioned above.
Another option is to use shell primitives such as \ttt{\&}, \ttt{wait}, \ttt{for} to explicitly induce concurrency;
 these come at a cost of manual effort to split inputs, rewrite scripts, and orchestrate execution---an expensive and error-prone process.
% re-writing of the entire script from scratch; unfortunately, this process is expensive and can introduce new bugs, cascading changes, and divergence from legacy functionality---but also foregoes all the productivity benefits of the original script.
% The fact that a programmer often belongs to both groups at the same time---as composed scripts can themselves form a command used in other scrips---compounds the challenge.
To top it off, all these options assume a good understanding of parallelism;
  users whose domain of expertise falls outside computing---routinely writing shell scripts to accomplish their tasks---are left without options (see~\S\ref{related} for a full discussion).
% \nv{I still don't like ``for a full discussion''---raises too many expectations, and we don't have that much more to say}

To address this challenge, we develop \sys, a shell that parallelizes POSIX shell scripts.
% \kk{Mention that it outputs shell scripts.}
% \nv{I think the ``how'' it solves it should come later, we're still trying to convince them we solve it (see next sentences in this paragraph)
% \sys offers both classes of users in the second group such much-needed automation.
\sys benefits both programmer groups mentioned earlier, with a particular emphasis on users.
Command developers are provided a set of abstractions, akin to lightweight type annotations, for expressing the parallelizability properties of their commands:
  rather than providing a command's full observable behavior, these annotations focus primarily on their interaction with state.
Shell users, on the other hand, are provided full automation:
  \sys extracts parallelism latent in their scripts by analyzing their parallelizability properties. % of their fragments.
% benefiting scripting experts and non-experts alike by offering automated POSIX-aware parallelization of shell-scripts as long as command developers
\sys's transformations are conservative, in that they do not attempt to parallelize fragments that lack sufficient information---\ie at worst, \sys will choose to not improve performance rather than risking breakage.
% \kk{Should we mention that we only show this empirically?}
% \nv{I don't think so; I don't think this is the right point to talk about our evaluation, we just state our intention/high-level principles.}

@@

These components are finally paired with \sys's runtime component.
% \sys's runtime  component additional packs
% Aside from these two components, \sys packs a runtime component that serves dual purpose:
Aware of the \unix philosophy and abstractions, it packs a small library of highly-optimized data combinators as well as high-performance primitives for eager data splitting and merging.
These address several practical challenges % such as task parallelism, deadlocks, and performance bottlenecks,
  and were developed by uncovering several pathological situations, on many of which we report.

@@

We evaluate \sys using
  (i) a series of benchmarks, ranging from classic one-liners found in the \unix literature to modern data-processing pipelines;
  (ii) three large and complex use cases, covering temperature analysis, web crawling and indexing, and container orchestration; and
  (iii) a series of targeted micro-benchmarks inspired by (and comparing with) possible alternatives.
The results demonstrate significant advantages in correctness, manual effort, and performance---with \sys's transformations offering an order-of-magnitude performance improvements aided further by \sys's optimized runtime components. \kk{I would remove the ``aided further...'' part.}
Naturally, these improvements depend on several factors such as the script's execution time, the pipeline depth, and the parallelizability class under which a command falls. 
%\kk{Do we need to mention this last sentence here? Without context it is pretty difficult for anyone to get something out of it.}

@@

The paper is structured as follows.
It starts by introducing the necessary background on shell scripting and overviewing \sys~\sx{bg}.
Sections \ref{distributability}--\ref{impl} highlight key contributions:
\begin{itemize}
  \item
  \S\ref{distributability}
    studies the parallelizability of a large set of shell commands and introduces a lightweight annotation language for commands that can be executed in a data-parallel manner.

@@

  \item
  \S\ref{impl} details \sys's implementation, discussing several challenges related to \sys's translation, optimization, and runtime phases.
% \kk{Rewrite that when we know for sure what the section contains.}

@@

\paragraph{Command Inputs and Outputs}

@@

\paragraph{Command-line Options}
As described earlier~\sx{bg}, com\-mand-line options and environment
variables can affect the operation of a command.  Options are a
particularly popular way to control a command, directly affecting a
command's parallelizability classification.  That is, commands are
assigned to default classes, and options can change the class in which
they belong.  As a simple example, consider \ttt{cat}.  By default
\ttt{cat} is in \sta, but with \ttt{-n} it jumps into \pur because it
has to keep track of a counter and output the line number with each line.

@@

%% \paragraph{Command Inputs and Outputs}
%% A second concern captured by the AL is the order of inputs and outputs.
%% Commands generally read multiple input streams (and, at cases, write to multiple output streams).
%% The invariants between inputs and outputs must be maintained over the distributed execution of the program.
%% % but these maintain different invariants, enabling very different distributability characteristics.

@@

\paragraph{Parallelizability Annotation Language}

@@

@@

\paragraph{Parallelizable Pure Commands}

@@

\paragraph{Edges -- Files}

@@

\paragraph{Nodes -- Commands}

@@

\paragraph{\todo{One-input-output} Commands}

@@

\paragraph{Stateless and Parallelizable Pure Commands}

@@

\paragraph{Parallelization Transformations}
%
Based on these equations, we can define a \emph{node parallelization
  transformation} on a stateless node $v$ that is preceded by a
\ttt{cat} with $n$ input streams and is followed by a node $v'$. The
transformation replaces $v$ with $n$ new nodes, routing each of the
$n$ input streams to one of them, and commutes the \ttt{cat} node
after them to concatenate their outputs and transfer them to
$v'$. Since each incoming edge represents a stream of data $x_i :
D^*$, and the only behaviour of a dataflow graph is its output, this
optimization $ v(x_1.x_2...x_n, s) \Rightarrow v(x_1, s).v(x_2,
s)...v(x_n, s)$ can be shown to preserve the behaviour of the
graph. Furthermore, it can be straightforwardly extended to pure
commands that can be implemented by a map-fold pair $(m, r)$ as $
v(x_1.x_2...x_n, s) \Rightarrow r(m(x_1, s).m(x_2, s)...m(x_n, s),
s)$. A visual representation of this transformation can be seen in
\Cref{fig:parallelization-transformation}.

@@

\paragraph{Parallelizable Regions}
%
%% \kk{The goal of this subsection is to explain that Dish does a safe
%%   but effective analysis to identify which parts of the scripts to
%%   parallelize.}
%
Parallelizable regions are program sub-expressions that can be
parallelized safely without altering the total program behaviour.
%% respect to their sequential execution.
The search for these regions can be guided by the structure of shells
programs, which inherently contains information about components that
can be executed independently and barriers that are natural points for
enforcing synchronization. Consider the following example:

@@

\paragraph{Translation Pass}

@@

\paragraph{Pure Fold Function Implementations}

@@

    \begin{subfigure}[b]{0.24\textwidth}
        \centering
        \includegraphics[width=\textwidth]{\detokenize{./figs/topn_throughput_scaleup.pdf}}
        %% \caption{TODO}
        %% \label{eval:topn}
    \end{subfigure}%
    ~
    \begin{subfigure}[b]{0.24\textwidth}
        \centering
        \includegraphics[width=\textwidth]{\detokenize{./figs/wf_throughput_scaleup.pdf}}
        %% \caption{TODO}
        %% \label{eval:wf}
    \end{subfigure}%
    ~
    \begin{subfigure}[b]{0.24\textwidth}
        \centering
        \includegraphics[width=\textwidth]{\detokenize{./figs/shortest_scripts_throughput_scaleup.pdf}}
        %% \caption{TODO}
        %% \label{eval:spell}
    \end{subfigure}%
    ~
    \begin{subfigure}[b]{0.24\textwidth}
      \centering
      \includegraphics[width=\textwidth]{\detokenize{./figs/alt_bigrams_throughput_scaleup.pdf}}
        %% \caption{TODO}
        %% \label{eval:minimal_sort}
    \end{subfigure}%
    \caption{
      \textbf{Speedup achieved by \sys as a function of the level of parallelism (1--100).} 
      Four different configurations per benchmark:
      (i) \todo{Parallel}, showing the performance of \sys with both eager and split optimizations enabled,
      (ii) \todo{No Split}, showing the performance of \sys with split disabled,
      (iii) \todo{No Eager}, showing the performance of \sys with split and eager disabled,
      (iv) \todo{Blocking Eager}, showing the performance of \sys with split disabled and the naive blocking eager enabled,
      \nv{Maybe it makes sense to show log x-axis?}
    }
    \vspace{-15pt}
    \label{fig:microbenchmarks}
\end{figure*}

\heading{Programs}
Tab.~\ref{tab:eval} summarizes the collection of micro-benchmarks used to evaluate \sys.
Benchmark names (col. 1) correspond to the names on the plots (Fig.~\ref{fig:microbenchmarks}), and structure (col. 2) summarizes the different classes of commands used in the script.
% The centralized pipelines read files of various sizes from disk and write to disk.
Input sizes and times (col. 3, 4) summarize the characteristics of the largest sequential executions. % (200 files in Fig.~\ref{fig:microbenchmarks}).
Script sizes (col. 5) report on \sys's rewritten output for three different distribution sizes---$1\times$, 10$\times$, and $100\times$.
Size grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
The first number alone is interesting, as it captures \sys's output for the sequential script.
As there is no parallelism involved, it highlights the initial (bare minimum) cost corresponding to using \sys.
% this is the size of 
This change in size does not necessarily translate to observed runtime, as most of the size increase comes from the setup and teardown of FIFOs responsible for interprocess (IPC) communication.
FIFOs are simply pipes made explicit;
  the performance characteristics, including in-kernel buffering and synchronization mechanisms, are identical between the two.

@@

\paragraph{Annotation-based Distribution}
Recent systems on light-touch distribution, such as
Ignis~\cite{ignis:19} and Mozart~\cite{mozart:19}, are more general
than cluster-computing frameworks.  They also lower (but not entirely
eliminate) developer effort by automating most of the rewriting
process.
% Both work on high-level, managed languages---server-side JavaScript for Ignis and Python for Mozart---that allow runtime transformations without affecting the broader environment under which a program executes.
For the critical points where user input is required, they provide a
high-level declarative DSL with varying (but limited) degrees of
expressive power.
% Annotations are used by ---complicated by the use of third-party
% libraries similar to \sys's extensions.
Unfortunately, a challenge with this approach is that users composing
applications by using third-party libraries do not understand their
distributability details well-enough to annotate them.
